---
title: "D10 Dashboard Project - Public Health Priorities for Scotland"
author: "D10 Data Solutions"
date: "09/09/2021"
output:
  ioslides_presentation:
        widescreen: true
        css: styles.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```
```{r, include=FALSE}
library(tidyverse)
library(here)
```
```{r, include=FALSE}
sdmd_combined_plus_zones <-
  read_csv(here("clean_data/sdmd_combined_plus_zones.csv"))

smoking <- read_csv(here("clean_data/smoking.csv"))

life_expectancy_clean <-
  read_csv(here("clean_data/life_expectancy.csv"))

color_theme <- function() {
  theme(

    plot.background = element_rect(fill ="white"),
    plot.title = element_text(size = rel(2)),
    plot.title.position = "plot",

    panel.border = element_rect(colour = "blue", fill = NA, linetype = 1),
    panel.background = element_rect(fill = "white"),
    panel.grid =  element_line(colour = "grey85", linetype = 1, size = 0.5),

    axis.text = element_text(colour = "blue", face = "italic", size = 14),
    axis.title.y = element_text(colour = "#1B732B" ,  size = 14, angle = 90),
    axis.title.x = element_text(colour = "#1B732B" ,  size = 14),


    legend.box.background = element_rect(),
    legend.box.margin = margin(6, 6, 6, 6)

  )
}

```




# Introduction | Jonathan

## The Team

- Derek - Life Expectancy SME
- John Paul - Smoking SME
- JP - Drug Abuse SME
- Prathiba - R Shiny Dashboard Architect
- Paddy - Spatial Data Guru





# Key findings - Life expectancy | Derek

## The Data

- What is life expectancy and why it was chosen?

Life expectancy is a statistical measure of the average time an organism is expected to live, based on the year of its birth, its current age and any other demographic factors including sex and geographic area.

- The source of the data

statistics.gov.uk

- Why was this data selected?

## Trend - Dumfries & Galloway

```{r, message=FALSE, fig.width=9,fig.height=5}
life_expectancy_clean %>%
  filter(type == "Local Authority",
         name == "Dumfries and Galloway",
         simd_quintiles == "All",
         urban_rural_classification == "All",
         sex != "All",
         age == 0,
         date_code %in% c("1991-1993", "1993-1995", "1995-1997", "1997-1999",
                   "1999-2001", "2001-2003", "2003-2005", "2005-2007",
                   "2007-2009", "2009-2011", "2011-2013", "2013-2015",
                   "2015-2017", "2017-2019")) %>%
  ggplot() +
  aes(x = date_code, y = le_value, group = sex, colour = sex) +
  geom_line() +
  geom_ribbon(aes(ymax = le_upper_ci, ymin = le_lower_ci), alpha = 0.25, colour = NA) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 0.5)) +
  labs(
    x = "\nYear",
    y = "Life Expectancy (years)\n",
    colour = "Sex"
  ) +
  ggtitle("Life Expectancy for Dumfries & Galloway over time\n") +
  color_theme()
```

## Trend - Scotland  

```{r,message=FALSE, fig.width=9,fig.height=5}
life_expectancy_clean %>%
  filter(type == "Scotland",
         simd_quintiles == "All",
         urban_rural_classification == "All",
         sex != "All",
         age == 0,
         date_code %in% c("1991-1993", "1993-1995", "1995-1997", "1997-1999",
                   "1999-2001", "2001-2003", "2003-2005", "2005-2007",
                   "2007-2009", "2009-2011", "2011-2013", "2013-2015",
                   "2015-2017", "2017-2019")) %>%
  ggplot() +
  aes(x = date_code, y = le_value, group = sex, colour = sex) +
  geom_line() +
  geom_ribbon(aes(ymax = le_upper_ci, ymin = le_lower_ci), alpha = 0.25, colour = NA) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 0.5)) +
  labs(
    x = "\nYear",
    y = "Life Expectancy (years)\n",
    colour = "Sex"
  ) +
  ggtitle("Life Expectancy for Scotland over time\n") +
  color_theme()
```
  

## Rank - Local Authority

```{r, message=FALSE, fig.width=9,fig.height=5}
life_expectancy_clean %>%
  filter(type == "Local Authority",
         simd_quintiles == "All",
         urban_rural_classification == "All",
         sex != "All",
         age == 0,
         date_code == "2017-2019") %>%
  ggplot() +
  aes(x = reorder(name, le_value), y = le_value, group = sex, colour = sex) +
  geom_line() +
  geom_ribbon(aes(ymax = le_upper_ci, ymin = le_lower_ci), alpha = 0.25, colour = NA) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  labs(
    title = "Life Expectancy by Local Authority",
    subtitle = "2017-19",
    x = "\nLocal Authority",
    y = "Life Expectancy (years)\n",
    colour = "Sex"
  )
```

## Key Findings

- In Scotland in 2017-2019, the life expectancy at birth was 77.1 years for males, and 81.1 years for females.
- Life expectancy since 1991-1993 has increased but has largely remained unchanged since 2013-2015.
- Female life expectancy is lowest at birth in Glasgow (78.5 years) and highest in East Renfrewshire (84 years).
- Male life expectancy is lowest at birth in Glasgow (73.6 years) and highest in East Dunbartonshire (80.5 years).
- The majority of Scotland's council areas have seen a slow down in life expectancy since 2013-2015, with some seeing a decrease.

# Key findings - Smoking | John Paul

## The data
- The data set used for this was:
Smoking - Scottish Survey Core Questions

- The data was sourced from:
statistics.gov.uk

- Why it was chosen
We chose Smoking as two of the Scottish Government's National Performance
Framework (NPF)  National Indicators are relevant to smoking.

There is a strong Public Health focus on reducing the proportion of adults who smoke.

## Trend
The overview is optimistic.  

Whilst numbers are still higher than anyone would like, the long term trend
shows a steady reduction in all age groups.

```{r, message=FALSE, warning=FALSE, fig.width=9,fig.height=4}
smoking %>%
  filter(smokes == "Yes",
         sex == "All") %>%

  arrange(date_code) %>%
  group_by(date_code, age) %>%
  summarise(avg_rate = mean(sm_percent)) %>%


  ggplot(aes(x = date_code, y = avg_rate, group = age, colour = age)) +
  geom_point() +
  geom_line() +
   theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
   labs(
    title = "Percentage of people who smoke",
    subtitle = "By age group",
    x = "\nYear",
    y = "Current Smokers (%)\n"
  ) +
  color_theme()
```

## Rank
In some areas you are 300% more likely to smoke than in other areas.

```{r, message=FALSE, warning=FALSE, fig.width=9,fig.height=5}
#  Highest Smoking rates per Local Authority area, in a 1 year period
smoking %>%
  filter(type == "Local Authority",
         type_of_tenure == "All",
         age == "All",
         household_type == "All",
         sex == "All",
         long_term_condition == "All",
         smokes == "Yes",
         date_code == "2019"
  ) %>%
 group_by(name, date_code, sm_percent) %>%
  select(name, smokes, sm_percent) %>%
  arrange(desc(sm_percent)) %>%
head() %>%
   ggplot(aes(x = reorder(name, sm_percent), y = sm_percent)) +
  geom_col() +
   theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust = 0.5)) +
   labs(
    title = "Highest Smoking rates per Local Authority area",
    subtitle = "2019",
    x = "\nLocal Authority",
    y = "Smoking 
    Rate (%)\n") +
  color_theme()
```

## Rank - continued
Where you live dictates how likely you are to smoke.

```{r, message=FALSE, warning=FALSE, fig.width=9,fig.height=4}
#  Lowest Smoking rates per Local Authority area, in a 1 year period
smoking %>%
  filter(type == "Local Authority",
         type_of_tenure == "All",
         age == "All",
         household_type == "All",
         sex == "All",
         long_term_condition == "All",
         smokes == "Yes",
         date_code == "2019"
  ) %>%
 group_by(name, date_code, sm_percent) %>%
  select(name, smokes, sm_percent) %>%
  arrange(sm_percent) %>%
  head() %>%
   ggplot(aes(x = reorder(name, sm_percent), y = sm_percent)) +
  geom_col() +
   theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust = 0.5)) +
   labs(
    title = "Lowest Smoking rates per Local Authority area",
    subtitle = "2019",
    x = "\nLocal Authority\n",
    y = "Smoking 
    Rate (%)\n") +
  color_theme()
```


## Map

There are huge variations around the country, whether they be measured by
NHS Health Board or the more granular Local Authority areas.   

Essentially the areas with the lowest proportion of smokers are more affluent. The
opposite is also true.


- Findings

More work needs to be in less affluent areas to further reduce the rate of smoking.


# Key findings - Drug abuse | Jonathan

## The data
- Scottish Drug Misuse Database
- https://www.opendata.nhs.scot/
- Published by Public Health Scotland - the national public health body for Scotland
- Why was this data selected?


## Trend by gender  

```{r, message=FALSE, fig.width=9,fig.height=5}
sdmd_combined_plus_zones %>%
  filter(sex != "All") %>%
  filter(number_assessed != 0) %>%
  filter(!is.na(number_assessed)) %>%
  filter(year != "2013/14") %>%
  arrange(year) %>%
  group_by(year, sex) %>%
  summarise(number_assessed_total = sum(number_assessed)) %>%
  ggplot(aes(x = year, y = number_assessed_total, group = sex, colour = sex)) +
  geom_point() +
  geom_line() +
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust = 1)) +
  labs(
    title = "Numbers assessed at specialist drug treatment centres over time",
    subtitle = "By gender\n",
    x = "\nNumber assessed\n",
    y = "\nYear\n",
  ) +
  color_theme()
```

## Trend by age group  

```{r, fig.width=9,fig.height=5}
sdmd_combined_plus_zones %>%
  filter(age != "All" &
         age != "Unknown" &
         year != "2013/14") %>%
  filter(number_assessed != 0) %>%
  filter(!is.na(number_assessed)) %>%
  arrange(year) %>%
  group_by(year, age) %>%
  summarise(number_assessed_total = sum(number_assessed), .groups = "drop") %>%
  ggplot(aes(x = year, y = number_assessed_total, group = age, colour = age)) +
  geom_point() +
  geom_line() +
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust = 1)) +
  labs(
    title = "Numbers assessed at specialist drug treatment centres over time",
    subtitle = "By age group\n",
    x = "\nYear\n",
    y = "\nNumber of people assessed\n"
    ) +
  color_theme()
```


## Rank  

```{r, message=FALSE, fig.width=9,fig.height=5}
sdmd_combined_plus_zones %>%
  filter(type == "NHS Health Board",
             year == "2018/19",
             sex == "All",
             name != "Shetland" &
             name != "Western Isles") %>%
      group_by(name) %>%
      summarise(total_number_assessed = sum(number_assessed)) %>%
      arrange(desc(total_number_assessed)) %>%
      ggplot() +
      aes(x = reorder(name, total_number_assessed), y = total_number_assessed) +
      geom_col() +
      theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
      labs(
        title = "Numbers assessed by area",
        subtitle = "NHS Health Boards\n",
        x = "Data Zone",
        y = "\nNumber
        assessed\n"
      )+
  color_theme()
```


# Dashboard Demonstration | Prathiba & Paddy


# Questions
