---
title: "Statistical Analysis for the public health data"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	message = FALSE,
	warning = FALSE
)
```


```{r}
#Load the library files
library(here)
library(tidyverse)
library(ggplot2)
library(RColorBrewer)
library(plotly)
library(lubridate)
```

```{r}
#Load the cleaning script
source (here("data_cleaning_wrangling/life_expectancy.R"))
```

```{r}
color_theme <- function() {
  theme(

    plot.background = element_rect(fill ="white"),
    plot.title = element_text(size = rel(2)),
    plot.title.position = "plot",

    panel.border = element_rect(colour = "blue", fill = NA, linetype = 1),
    panel.background = element_rect(fill = "white"),
    panel.grid =  element_line(colour = "grey85", linetype = 1, size = 0.5),

    axis.text = element_text(colour = "blue", face = "italic", size = 14),
    axis.title.y = element_text(colour = "#1B732B" ,  size = 14, angle = 90),
    axis.title.x = element_text(colour = "#1B732B" ,  size = 14),


    legend.box.background = element_rect(),
    legend.box.margin = margin(6, 6, 6, 6)

  )
}
```

```{r}
#Load the cleaned data
life_expectancy <- read_csv(here("clean_data/life_expectancy.csv"))
```

# Life Expectancy

### Plot 1: Life expectancy for Scotland over time (Trend)

```{r}

# this could be coded to allow the user to select age & urban_rural_classification
p <- life_expectancy_clean %>% 
  filter(type == "Scotland",
         age == 0,
         simd_quintiles == "All",
         urban_rural_classification == "All") %>%
  arrange(date_code) %>% 
  ggplot() +
  aes(x = date_code, y = le_value, group = sex, colour = sex) +
  geom_line() +
  geom_point() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  labs(
    x = "\nYear",
    y = "Life Expectancy (years)\n",
    colour = "Sex"
  )+
  color_theme()

ggplotly(p)
```

### Plot 2a: Comparison of life expectancy for on NHS Health Boards (Rank - Top 5)

```{r}
p <- life_expectancy_clean %>% 
  filter(type == "NHS Health Board",
         age == 0,
         sex == "All",
         simd_quintiles == "All",
         urban_rural_classification == "All",
         date_code == "2013-2015") %>%
  arrange(desc(le_value)) %>% 
  head(n = 5) %>% 
  ggplot() +
  aes(x = reorder(name, le_value), y = le_value, fill = le_value) +
  geom_col(position = "dodge") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  labs(
    x = "\nLocal authority",
    y = "Life Epectancy (years)\n",
    fill = "Sex"
  )+
  color_theme()+
  scale_fill_distiller(palette = "YlGn")

ggplotly(p)
```

### Plot 2b: Comparison of life expectancy for all local authorities (Rank - Lowest 5)

```{r}
# this could be coded to allow the user to select local authority or NHS health board, date_code & age
p <- life_expectancy_clean %>% 
  filter(type == "Local Authority",
         age == 0,
         sex == "All",
         simd_quintiles == "All",
         urban_rural_classification == "All",
         date_code == "2013-2015") %>%
  arrange(desc(le_value)) %>% 
  tail(n = 5) %>%
  ggplot() +
  aes(x =  reorder(name, le_value), y = le_value, fill = le_value) +
  geom_col(position = "dodge") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  labs(
    x = "Local authority",
    y = "Life Epectancy (years)\n",
    fill = "Sex"
  )+
  color_theme()+
  scale_fill_distiller(palette = "YlOrRd")

ggplotly(p)
```

# Smoking Analysis

```{r}
# Load the cleaned data
smoking <- read_csv(here("clean_data/smoking.csv")) %>% 
  mutate(entry_id = row_number())
```

### Plot 1: Countrywide average smoking rates over time (in %) (Trend)

```{r}
p <- smoking %>%
  filter(smokes == "Yes",
         sex == "All") %>%
  
  arrange(date_code) %>% 
  group_by(date_code, age) %>% 
  summarise(avg_rate = mean(sm_percent)) %>% 
  

  ggplot(aes(x = date_code, y = avg_rate, group = age, colour = age)) +
  geom_point() +
  geom_line() +
   theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
   labs(
    x = "\nYear",
    y = "Current Smokers (%)\n"
)+
  color_theme()

ggplotly(p)

```

### Plot 2a: Smoking rates per Local Authority area, in a 1 year period (Rank - Highest)

```{r}
#  Highest Smoking rates per Local Authority area, in a 1 year period
p <- smoking %>% 
  filter(type == "Local Authority",
         type_of_tenure == "All",
         age == "All",
         household_type == "All",
         sex == "All",
         long_term_condition == "All",
         smokes == "Yes",
         date_code == "2019"
  ) %>%
 group_by(name, sm_percent) %>% 
  select(name, smokes, sm_percent) %>% 
  arrange(desc(sm_percent)) %>% 
head() %>% 
   ggplot(aes(x = reorder(name, sm_percent), y = sm_percent, fill = sm_percent)) +
  geom_col() +
   theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
   labs(
    x = "\nLocal Authority",
    y = "2019 Smoking Rate (%)\n")+
  scale_fill_distiller(palette = "YlGn")+
  color_theme()

ggplotly(p)
```

### Plot 2b: Smoking rates per Local Authority area, in a 1 year period (Rank - Lowest)

```{r}
#  Lowest Smoking rates per Local Authority area, in a 1 year period
p <- smoking %>% 
  filter(type == "Local Authority",
         type_of_tenure == "All",
         age == "All",
         household_type == "All",
         sex == "All",
         long_term_condition == "All",
         smokes == "Yes",
         date_code == "2019"
  ) %>%
 group_by(name, sm_percent) %>% 
  select(name, smokes, sm_percent) %>% 
  arrange(sm_percent) %>% 
  head() %>% 
   ggplot(aes(x = reorder(name, sm_percent), y = sm_percent, fill = sm_percent)) +
  geom_col() +
   theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
   labs(
    x = "\nLocal Authority",
    y = "2019 Smoking Rate (%)\n")  +
  scale_fill_distiller(palette = "YlOrRd")+
  color_theme()

ggplotly(p)
```

# Drug Abuse Analysis



```{r}
#Read-in cleaned data
drugs <- read_csv(here("clean_data/sdmd_combined_plus_zones.csv"))

```

```{r include=FALSE}
#Explore/investigate
str(drugs)
```

### Plot 1a: Drug abuse over time by number assessed, split by gender (Trend)

```{r}
p <- drugs %>% 
  filter(sex != "All") %>% 
  filter(number_assessed != 0) %>%
  filter(!is.na(number_assessed)) %>%
  filter(year != "2013/14") %>% 
  arrange(year) %>% 
  group_by(year, sex) %>% 
  summarise(number_assessed_total = sum(number_assessed)) %>% 
  ggplot(aes(x = year, y = number_assessed_total, group = sex, colour = sex)) +
  geom_point() +
  geom_line()+
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust = 1)) +
  color_theme()+
   labs(
    x = "\nYear",
    y = "Number Assessed\n") 

ggplotly(p)
```

### Plot 1b: Drug abuse over time by number assessed, split by age group (Trend)

```{r}
p <- drugs %>% 
  filter(age != "All" &
         age != "Unknown" &
         year != "2013/14") %>% 
  filter(number_assessed != 0) %>%
  filter(!is.na(number_assessed)) %>% 
  arrange(year) %>% 
  group_by(year, age) %>% 
  summarise(number_assessed_total = sum(number_assessed), .groups = "drop") %>% 
  ggplot(aes(x = year, y = number_assessed_total, group = age, colour = age)) +
  geom_point() +
  geom_line()+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  color_theme()+
   labs(
    x = "\nYear",
    y = "Number Assessed\n") 

ggplotly(p)
  
```

### Plot 2a: Drug abuse by area (average for Scotland could also be added) (Rank - Highest)

```{r}

p <- drugs %>% 
  filter(type == "Local Authority" &
         sex == "All" &
         # age_group ==   &
         number_assessed != 0 &
         !is.na(number_assessed)) %>% 
  group_by(name) %>% 
  summarise(number_assessed_total = sum(number_assessed)) %>% 
  arrange (desc(number_assessed_total)) %>% 
  head(n=5) %>% 
  ggplot(aes(x = reorder(name,number_assessed_total), y = number_assessed_total, fill = number_assessed_total)) +
  geom_col() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))+
  color_theme() +
  scale_fill_distiller(palette = "YlGn")+
   labs(
    x = "\nLocal Authority",
    y = "Number Assessed\n") 

ggplotly(p)
  
```

### Plot 2: Drug abuse by area (average for Scotland could also be added) (Rank - Lowest)

```{r}

p <- drugs %>% 
  filter(type == "Local Authority" &
         sex == "All" &
         # age_group ==   &
         number_assessed != 0 &
         !is.na(number_assessed)) %>% 
  group_by(name) %>% 
  summarise(number_assessed_total = sum(number_assessed)) %>% 
  arrange (desc(number_assessed_total)) %>% 
  tail(n=5) %>% 
  ggplot(aes(x = reorder(name,number_assessed_total), y = number_assessed_total, fill = number_assessed_total)) +
  geom_col() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))+
  color_theme() +
  scale_fill_distiller(palette = "YlOrRd")+
   labs(
    x = "\nLocal Authority",
    y = "Number Assessed\n") 

ggplotly(p)
  
```
